---
title: "GCIMS R Package case study"
author: "S. Oller-Moreno*, C. Mallafré-Muro*, L. Fernández, E. Caballero, A. Blanco,  J. Gumà, A. Pardo, S. Marco"
date: "`r format(Sys.Date(), '%F')`"
abstract: >
  An implementation of the GCIMS package, showing the most relevant functions and
  a proposed workflow. This includes urine samples, adding sample
  annotations, preprocessing the spectra, alignment, detecting peaks and regions
  of interest (ROIs), clustering of ROIs across samples, peak integration and
  building a peak table.
vignette: >
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
start_time <- Sys.time()
library(BiocParallel)
library(ggplot2)
library(GCIMS)
```


The GCIMS package allows you to import your Gas Chromatography - Ion Mobility Spectrometry samples,
preprocess them, align them one to each other and build a peak table with the relevant features.

Enable parallelization of the workflow:

```{r}
show_progress_bar <- interactive() && is.null(getOption("knitr.in.progress"))

# disable parallellization: (Useful for better error reporting)
#register(SerialParam(progressbar = show_progress_bar), default = TRUE)

# enable parallellization:
workers<-parallel::detectCores()-4
register(SnowParam(workers = workers, progressbar = show_progress_bar, exportglobals = FALSE), default = TRUE)
```

# Downloading the dataset

The tutorial of the GCIMS R Package can be tested with a dataset available in Zenodo.
This dataset includes a set of urine samples measured with a GC-IMS FlavourSpec.

A .zip file with the urines will be downloaded in a temporal file and extracted in a new folder in the working directory
```{r}
samples_directory <- file.path(getwd(), "Urines")
dir.create(samples_directory)
tmp_zipfile<-tempfile(fileext = ".zip")
curl::curl_download(url = "https://zenodo.org/record/7941230/files/Urines.zip?download=1", destfile = tmp_zipfile, quiet = FALSE)
utils::unzip(tmp_zipfile,junkpaths = TRUE, exdir = samples_directory)
```

Other files needed to run this vignette will also be downloaded in the folder created

```{r}
curl::curl_download(url = "https://zenodo.org/record/7941230/files/annotations.csv?download=1", 
                    destfile = file.path(samples_directory,"annotations.csv"), quiet = FALSE)
curl::curl_download(url = "https://zenodo.org/record/7941230/files/reference_peaks.csv?download=1", 
                    destfile = file.path(samples_directory,"reference_peaks.csv"), quiet = FALSE)

```
Check that the files are downloaded:

```{r}
list.files(samples_directory)
```

# Import data

The next step is to import all the samples information and maintain only the urines information. Remove the samples information which do not correspond to urines. 

```{r message=FALSE}
annotations <- readr::read_csv(file.path(samples_directory, "annotations.csv"))
annotations
```


# Create a GCIMSDataset object

The GCIMS package woks with GCIMSDataset objects, which needs to be created.

```{r}
dataset <- GCIMSDataset$new(
  annotations,
  base_dir = getwd(),
  on_ram = FALSE)
dataset
```


Most operations on the `dataset` are not executed until you need to get the actual samples or
data. This is done to perform them in batch, more efficiently, if possible. However,
you can manually `realize` the `GCIMSDataset` object so it executes all its pending operations.
We can see how the "read_sample" pending operation becomes part of the dataset history:

```{r}
dataset$realize()
dataset
```


```{r}
Sample_1 <- dataset$getSample(sample = "220221_151456")

dt_k1 <- dtime(Sample_1)
tis_k1 <- getTIS(Sample_1)
ggplot(dplyr::filter(data.frame(x = dt_k1, y = tis_k1), x >1)) + 
  geom_line(aes(x = x, y = y)) +
  scale_x_continuous(name = "Drift time (ms)", limits = c(7, 17)) +
  scale_y_continuous(name = "Intensity (a.u)", trans = cubic_root_trans())

rt_k1 <- rtime(Sample_1)
ric_k1 <- getRIC(Sample_1)

ggplot(dplyr::filter(data.frame(x = rt_k1, y = ric_k1))) + 
  geom_line(aes(x = x, y = y)) +
  scale_x_continuous(name = "Retention time (ms)", limits = c(55, 900)) +
  scale_y_continuous(name = "Intensity (a.u)")
```


# Filter the retention and drift time of your samples - Smoothing 

You can remove noise from your sample using a Savitzky-Golay filter, applied
both in drift time and in retention time.

The Savitzky-Golay has two main parameters: the filter length and the filter order.
It is recommended to use a filter order of 2, but the filter length must be selected
so it is large enough to remove noise but always smaller than the peak width to
prevent distorting the peaks.

You can apply the smoothing filter to a single IMS spectrum or to a single chromatogram
to see how noise is removed and how peaks are not distorted. Tweak the filter lengths
and, once you are happy, apply the smoothing filter to all the dataset.

```{r}
filterRt(dataset, rt = c(0, 1300)) # in s
filterDt(dataset, dt = c(7, 17)) # in ms
dataset
```

```{r}
ket1afterfilter <- dataset$getSample(sample = "220221_151456")
ket1_smoothed <- smooth(
  ket1afterfilter,
  rt_length_s = 3,
  dt_length_ms = 0.14,
  rt_order = 2,
  dt_order = 2
)
```


```{r}
dataset <- smooth(dataset, rt_length_s = 3, dt_length_ms = 0.14)
dataset$realize()
```

# Decimation

```{r}
ket1_decimated <- decimate(ket1_smoothed, rt_factor = 1, dt_factor = 2)

dataset <- decimate(dataset, rt_factor = 1, dt_factor = 2)
```

# Alignment

Pressure and temperature fluctuations as well as degradation of the chromatographic
column are some of the causes of missalignments in the data, both in retention and
drift time.

In order to be able to compare samples to each other, we align the samples.

The alignment will happen only in drift time. To correct the drift time, we will use 
a Picewise multiplicative correction $t_d' = k  t_d$ according to a set of reference 
peaks. This reference peaks are present in all the samples and are previously manually
annotated.


```{r}
plotTIS(dataset, dt_range = c(7, 17))
```

```{r}
plotRIC(dataset)
```

```{r alignment-starts-here}
# Piecewise Multiplicative Correction

manual_align <- function(object, referencepeaks) {
  tramo1 <- referencepeaks[which(referencepeaks$Peak == "Injection Point"), ]
  ref_tramo1 <- apply(tramo1[ ,c(3, 4)], 2, median)

  fit_tramo1 <- NULL
  for(i in c(1:dim(tramo1)[1])){
    a <- rbind(c(0, 1), c(as.numeric(tramo1[i, 3]), 1))
    b <- c(0, ref_tramo1[1])
    fit_tramo1 <- rbind(fit_tramo1, solve(a, b))
  }
  
  corrected_tramo1 <- as.vector(fit_tramo1[,1]) * as.vector(tramo1[, 3]$RetentionTime_s) + fit_tramo1[,2]
  
  tramo2 <- referencepeaks[which(referencepeaks$Peak == 70), ]
  ref_tramo2 <- apply(tramo2[ ,c(3, 4)], 2, median)
  
  fit_tramo2 <- NULL
  for(i in c(1:dim(tramo2)[1])){
    a <- rbind(c(as.numeric(tramo1[i, 3]), 1), c(as.numeric(tramo2[i, 3]), 1))
    b <- c(ref_tramo1[1], ref_tramo2[1])
    fit_tramo2 <- rbind(fit_tramo2, solve(a, b))
  }
    
  corrected_tramo2 <- as.vector(fit_tramo2[,1]) * as.vector(tramo2[, 3]$RetentionTime_s) + fit_tramo2[,2]
    
  tramo3 <- referencepeaks[which(referencepeaks$Peak == 200), ]
  ref_tramo3 <- apply(tramo3[ ,c(3, 4)], 2, median)
    
  fit_tramo3 <- NULL
  for(i in c(1:dim(tramo3)[1])){
    a <- rbind(c(as.numeric(tramo2[i, 3]), 1), c(as.numeric(tramo3[i, 3]), 1))
    b <- c(ref_tramo2[1], ref_tramo3[1])
    fit_tramo3 <- rbind(fit_tramo3, solve(a, b))
  }
    
  corrected_tramo3 <- as.vector(fit_tramo3[,1]) * as.vector(tramo3[, 3]$RetentionTime_s) + fit_tramo3[,2]
    
  tramo4 <- referencepeaks[which(referencepeaks$Peak == 600), ]
  ref_tramo4 <- apply(tramo4[ ,c(3, 4)], 2, median)
    
  fit_tramo4 <- NULL
  for(i in c(1:dim(tramo4)[1])){
    a <- rbind(c(as.numeric(tramo3[i, 3]), 1), c(as.numeric(tramo4[i, 3]), 1))
    b <- c(ref_tramo3[1], ref_tramo4[1])
    fit_tramo4 <- rbind(fit_tramo4, solve(a, b))
  }
    
  corrected_tramo4 <- as.vector(fit_tramo4[,1]) * as.vector(tramo4[, 3]$RetentionTime_s) + fit_tramo4[,2]
    
  og_datos <- c(as.vector(tramo1[, 3]$RetentionTime_s), 
                as.vector(tramo2[, 3]$RetentionTime_s), 
                as.vector(tramo3[, 3]$RetentionTime_s), 
                as.vector(tramo4[, 3]$RetentionTime_s))
  corrected_datos <- c(corrected_tramo1, corrected_tramo2, corrected_tramo3, corrected_tramo4)
  plot(og_datos, corrected_datos, type = "b")
    
  #wd <- GCIMS:::CurrentHashedDir(object)
  d<-list()
  for(i in c(1:length(object$pData$SampleID))){
    #sample2align <- readRDS(paste0(wd, "/",annotations$SampleID[i], ".rds"))
    sample2align<-object$getSample(annotations$SampleID[i])
    int_mat <- intensity(sample2align)
    ret_time <- rtime(sample2align)

    rt_final <- seq(from = 0, to = ref_tramo4[1], by = 0.39)
  
    og_ancla1 <- which(round(ret_time, digits = 0) == as.numeric(tramo1[i, 3]))[2]
    og_ancla2 <- which(round(ret_time, digits = 0) == as.numeric(tramo2[i, 3]))[2]       
    og_ancla3 <- which(round(ret_time, digits = 0) == as.numeric(tramo3[i, 3]))[2]
    og_ancla4 <- which(round(ret_time, digits = 0) == as.numeric(tramo4[i, 3]))[2]

    rt_corr1 <- ret_time[1 : og_ancla1]* fit_tramo1[i,1] + fit_tramo1[i,2]
    rt_corr2 <- ret_time[(og_ancla1 + 1) : og_ancla2]* fit_tramo2[i,1] + fit_tramo2[i,2]
    rt_corr3 <- ret_time[(og_ancla2 + 1) : og_ancla3]* fit_tramo3[i,1] + fit_tramo3[i,2]
    rt_corr4 <- ret_time[(og_ancla3 + 1) : og_ancla4]* fit_tramo4[i,1] + fit_tramo4[i,2]
    
    rt_corr <- c(rt_corr1, rt_corr2, rt_corr3, rt_corr4)
    int_corr <- t(apply(int_mat, 1, function(y) signal::interp1(rt_corr, y, rt_final)))
    int_corr[is.na(int_corr)] <- int_mat[which(is.na(int_corr))]
    ncol(int_mat) == length(ret_time)
    ncol(int_corr) == length(rt_final)
    sample2align@retention_time <- rt_final
    sample2align@data <- int_corr
    #saveRDS(sample2align, paste0(wd, "/",annotations$SampleID[i], ".rds"))
    d[[annotations$SampleID[i]]]<-sample2align
  }
  return(d)
}
referencepeaks <- readr::read_csv(file.path(samples_directory, "reference_peaks.csv"))
referencepeaks$DriftTime_ms <- as.numeric(referencepeaks$DriftTime_ms)
d<-manual_align(dataset, referencepeaks)
dataset<-GCIMSDataset_fromList(d,on_ram=FALSE)
```

```{r}
dataset$realize()
```

```{r}
plotTIS(dataset, dt_range = c(7, 17))
```

```{r}
plotRIC(dataset, rt_range = c(50, 230))

plotRIC(dataset)

plotRIC(dataset, rt_range = c(180, 205))
```


# Peaks

First try one sample and optimize the `dt_peakwidth_range_ms` and `rt_peakwidth_range_s` parameters there. 
Change values according to the width your peaks present in the plots. 

```{r}
ket1 <- dataset$getSample(sample = "220221_151456")
ket1ensenal <- findPeaks(
  ket1,
  rt_length_s = 3,
  dt_length_ms = 0.14,
  verbose = TRUE,
  dt_peakwidth_range_ms = c(0.1, 0.4),
  rt_peakwidth_range_s = c(5, 25),
  dt_peakDetectionCWTParams = list(SNR.Th = 2, exclude0scaleAmpThresh = TRUE),
  rt_peakDetectionCWTParams = list(SNR.Th = 2, exclude0scaleAmpThresh = TRUE),
  dt_extension_factor = 0,
  rt_extension_factor = 0,
  exclude_rip = TRUE,
  iou_overlap_threshold = 0.2,
  debug_idx = list(rt = 204, dt = 1167)
)
peak_list_ket1 <- peaks(ket1ensenal)
plot(ket1ensenal, dt_range = c(9.2, 10), rt_range = c(50, 90)) + overlay_peaklist(peak_list_ket1, color_by = "PeakID")
```

```{r}
findPeaks(
  dataset,
  rt_length_s = 3,
  dt_length_ms = 0.14,
  verbose = TRUE,
  dt_peakwidth_range_ms = c(0.15, 0.4),
  rt_peakwidth_range_s = c(10, 25),
  dt_peakDetectionCWTParams = list(exclude0scaleAmpThresh = TRUE),
  rt_peakDetectionCWTParams = list(exclude0scaleAmpThresh = TRUE),
  dt_extension_factor = 0,
  rt_extension_factor = 0,
  exclude_rip = TRUE,
  iou_overlap_threshold = 0.2
)
peak_list <- peaks(dataset)
head(peak_list)
```

Plot all the peaks from all the dataset together, overlayed on a single sample:

```{r}
Sample_1 <- dataset$getSample(sample = "220221_151456")
plot(Sample_1, dt_range = c(7.5, 10), rt_range =  c(50, 200)) +overlay_peaklist(peaks(dataset), color_by = "SampleID")
```



# Clustering

```{r clustering-section-starts-here}
peak_clustering <- clusterPeaks(
  peak_list,
  distance_method = "euclidean",
  dt_cluster_spread_ms = 0.1,
  rt_cluster_spread_s = 20,
  distance_between_peaks_from_same_sample = 100,
  clustering = list(method = "hclust")
)

#plot(Sample_1)+ overlay_peaklist(peak_clustering$cluster_stats, color_by = "cluster")
```


The peak list, with cluster ids can be plotted on top of a single sample:

```{r}
peak_list_clustered <- peak_clustering$peak_list_clustered

tt <- merge(peak_list_clustered, annotations, by = "SampleID") #peak_list_clustered
Sample_1 <- dataset$getSample( sample = "220221_151456")
plot(Sample_1) +overlay_peaklist(tt, color_by = "cluster") + theme(legend.position = "none")
```


FIXME: We need a proper API to plot this with a user friendly interface, but we
can easily see there is some trouble with the clustering as of today (I'm working on it).

```{r eval=FALSE, fig.height=10, fig.width=10, include=FALSE}
plot(Sample_1) +overlay_peaklist(peak_list_clustered, 
                                        color_by = "SampleID") +
  lims(x = c(NA_real_, NA_real_), 
      y = c(NA_real_, NA_real_)) +
  facet_wrap(~cluster, scales = "free")
```

The resulting cluster sizes (median position of individual clusters) is not a good reference
for integration. We are working on this.

```{r}
plot(Sample_1) + overlay_peaklist(peak_clustering$cluster_stats, color_by = "cluster")
```

# Baseline correction

```{r}
dataset <- estimateBaseline(
  dataset,
  dt_peak_fwhm_ms = 0.2, 
  dt_region_multiplier = 12,
  rt_length_s = 200
)
dataset$realize()
```


```{r eval=FALSE, fig.height=10, fig.width=10, include=FALSE}
plt <- plot(Sample_1) + overlay_peaklist(dplyr::filter(peak_clustering$peak_list_clustered),
  color_by = "SampleID",
  col_prefix = "fixedsize_"
) + 
  lims(x = c(NA_real_, NA_real_), y = c(NA_real_, NA_real_)) +
  facet_wrap(~cluster, scales = "free")
plt
```


# Peak integration

```{r}
dataset <- integratePeaks(
  dataset, 
  peak_clustering$peak_list, 
  integration_size_method = "fixed_size", 
  rip_saturation_threshold = 0.1
)

```

```{r}
peak_list1 <- peaks(dataset)
```


# Build peak table

```{r}
peak_table <- peakTable(peak_list1, aggregate_conflicting_peaks = max)
peak_table$peak_table_matrix[1:10,1:10]
```

# Imputation

```{r}
peak_table_imputed <- imputePeakTable(peak_table$peak_table_matrix, dataset, peak_clustering$cluster_stats)
peak_table_imputed[1:10,1:10]
```

#Normalization 
```{r }
norm_pqn <- function(spectra) {
num_samples <- nrow(spectra)
if (num_samples < 10) {
  rlang::warn(message = c("There are not enough samples for reliably estimating the median spectra", "i" = paste0("The Probabalistic Quotient Normalization requires several samples ", "to compute the median spectra. Your number of samples is low"), "i" = paste0("Review your peaks before and after normalization to ","ensure there are no big distortions")))
}
# Normalize to the area
areas <- rowSums(spectra)
areas <- areas / stats::median(areas)
if (num_samples == 1) {
  # We have warned, and here there is nothing to do anymore
  rlang::warn("PQN is meaningless with a single sample. We have normalized it to the area.")
  out <- list(spectra = spectra / areas, norm_factor = areas)
  return(out)
}
spectra2 <- spectra / areas
# Move spectra above zero:
if (any(spectra2 <= 0)) {
  spectra2 <- spectra2 - min(spectra2)
}
# Median of each ppm: (We need multiple spectra in order to get a reliable median!)
m <- matrixStats::colMedians(as.matrix(spectra2))
# Divide at each ppm by its median:
f <- spectra2 / m[col(spectra2)]
f[which(is.na(f) == TRUE)] <- 0
if (any(f <= 0)) {
  f <- f - min(f)
}
f <- matrixStats::rowMedians(as.matrix(f))
# Divide each spectra by its f value
out <- list(spectra = spectra / (f * areas), norm_factor = f * areas)
out
}

peak_table_imputed_normalized<-norm_pqn(as.matrix(peak_table_imputed[,-which(colnames(peak_table_imputed)=="NA")]))
write.csv(peak_table_imputed_normalized$spectra,file.path(samples_directory,"peak_table_R.csv"))
```


```{r PLSDA}
dataR<-as.matrix(peak_table_imputed_normalized$spectra)
id<-annotations$ID
labels<-as.numeric(annotations$Sex=="Male")
pred <- data.frame()
lvh<-list()
for (i in unique(id)) {
  Xtrain <- dataR[id!=i, ]
  Ytrain<-labels[id!=i]
  idL<-id[id!=i]
  predL<-list()
  for (k in unique(idL)){
    XtrainL<-Xtrain[idL!=k,]
    YtrainL<-Ytrain[idL!=k]
    XtestL<-as.matrix(Xtrain[idL==k,])
    YtestL<-Ytrain[idL==k]
    if (length(XtestL[,1])!=length(YtestL)){
      XtestL<-t(XtestL)
    }
    plsL<-pls::plsr(YtrainL ~ XtrainL, ncomp=10, scale=TRUE)
    pL<-predict(plsL, XtestL,ncomp = 1:10)
    predL$a<-rbind(predL$a,cbind(pL[,,1],YtestL))
    predL$b<-rbind(predL$b,cbind(pL[,,2],YtestL))
    predL$c<-rbind(predL$c,cbind(pL[,,3],YtestL))
    predL$d<-rbind(predL$d,cbind(pL[,,4],YtestL))
    predL$e<-rbind(predL$e,cbind(pL[,,5],YtestL))
    predL$f<-rbind(predL$f,cbind(pL[,,6],YtestL))
    predL$g<-rbind(predL$g,cbind(pL[,,7],YtestL))
    predL$h<-rbind(predL$h,cbind(pL[,,8],YtestL))
    predL$l<-rbind(predL$l,cbind(pL[,,9],YtestL))
    predL$m<-rbind(predL$m,cbind(pL[,,10],YtestL))
  }
  aucL<-c(pROC::auc(response=predL$a[,2],predictor=predL$a[,1]),
          pROC::auc(response=predL$b[,2],predictor=predL$b[,1]),
          pROC::auc(response=predL$c[,2],predictor=predL$c[,1]),
          pROC::auc(response=predL$d[,2],predictor=predL$d[,1]),
          pROC::auc(response=predL$e[,2],predictor=predL$e[,1]),
          pROC::auc(response=predL$f[,2],predictor=predL$f[,1]),
          pROC::auc(response=predL$g[,2],predictor=predL$g[,1]),
          pROC::auc(response=predL$h[,2],predictor=predL$h[,1]),
          pROC::auc(response=predL$l[,2],predictor=predL$l[,1]),
          pROC::auc(response=predL$m[,2],predictor=predL$m[,1]))
  LV<-which.max(aucL)
  lvh <- c(lvh, LV)
  Xtrain <- dataR[id!=i,]
  Ytrain <- labels[id!=i]
  Xtest <- as.matrix(dataR[id==i,])
  Ytest <- labels[id==i]
  if (length(Xtest[,1])!=length(Ytest)){
    Xtest<-t(Xtest)
  }
  
  pls <- pls::plsr(Ytrain ~ Xtrain, ncomp=LV, scale=TRUE)
  p<-predict(pls, Xtest,ncomp = LV)
  pred<-rbind(pred,cbind(p[,,],Ytest))
}

AUC<-pROC::auc(response=pred$Ytest,predictor=pred$V1)

```

#Session Info:
```{r }
sessionInfo()
```